---
title: "STAT 545A Homework 3"
author: "Junbin ZHANG"
date: "Sept 25, 2018"
output:
  html_document:
    keep_md: true
    toc: true
    theme: readable
---

# Bring rectangular data in
```{r}
## load gapminder
suppressPackageStartupMessages(library(gapminder))
## load tidyverse
suppressPackageStartupMessages(library(tidyverse))
```

# Tasks

## Task #1

**Get the maximum and minimum of GDP per capita for all continents.**

We first create a table to show the maximum and minimum of GDP per capita for all continents.

```{r}
# get data of maximum and minimum of GDP per capita for all continents
data <- gapminder %>% 
  # group by continent
  group_by(continent) %>% 
  # calculate summaries
  summarize(
    max_gdpPercap = max(gdpPercap),
    min_gdpPercap = min(gdpPercap),
  )
# get corresponding rows for the maximum data
max_country <- gapminder %>% 
  # filter data with maximum GDP per capita
  filter(gdpPercap %in% data$max_gdpPercap) %>% 
  # create column with country and year
  mutate(
    max_country_year = paste(country, year, sep=",")
  ) %>% 
  # make sure the order is the same as "data"
  arrange(continent) %>% 
  # select only mutated column
  select(max_country_year)

# get corresponding rows for the minimum data
min_country <- gapminder %>% 
  # filter data with minimum GDP per capita
  filter(gdpPercap %in% data$min_gdpPercap) %>% 
  # create column with country and year
  mutate(
    min_country_year = paste(country, year, sep=",")
  ) %>% 
  # make sure the order is the same as "data"
  arrange(continent) %>% 
  # select only mutated column
  select(min_country_year)

# display table with combined data
data.frame(
  continent = data$continent,
  max_gdpPercap = data$max_gdpPercap,
  max_country,
  min_gdpPercap = data$min_gdpPercap,
  min_country
) %>% 
  knitr::kable()
```

Let's show them on a scatterplot.

```{r}
gapminder %>%
  # create a new column to indicate whether a data point is maximum or minimum gdpPercap
  mutate(
    max_min_gdpPercap = ifelse(
      gdpPercap %in% data$max_gdpPercap,
      "Maximum",
      ifelse(
        gdpPercap %in% data$min_gdpPercap,
        "Minimum",
        "Others"
        )
      )
  ) %>% 
  # year as x axis and gdpPercap as y axis, use max_min_gdpPercap as color
  ggplot(aes(year, gdpPercap, color = max_min_gdpPercap), alpha = 0.1) +
  # make it a scatterplot
  geom_point() +
  # facetting using continent
  facet_wrap(~continent, scales = "free_y") +
  # make a better x axis
  scale_x_continuous(breaks = scales::pretty_breaks(n = 3))
```

From the table and figure, we notice the maximum and minimum values of gdpPercap are different in different continents. They also appear in different countries and years. The scatterplot provides a better view, except we cannot get which countries has the maximum/minimum gdpPercap.

## Task #2

**Look at the spread of GDP per capita within the continents.**

Let's show a table with the range, mean and deviation of GDP per capita within the continents.

```{r}
gapminder %>%
  # group by continent
  group_by(continent) %>%
  # calculate summaries
  summarize(
    # calculate minimum
    min_gdpPercap = min(gdpPercap),
    # calculate maximum
    max_gdpPercap = min(gdpPercap),
    # calculate mean
    mean_gdpPercap = mean(gdpPercap),
    # calculate standard deviation
    sd_gdpPercap = sd(gdpPercap)
  ) %>% 
  # display the table
  knitr::kable()
```


We can use histogram and boxplot to do the same task. Notice that in histogram, `scales="free"` is needed to scale both x and y axes, so the figure should be clearer. 

```{r}
gapminder %>% 
  # gdpPercap as x axis
  ggplot(aes(gdpPercap, color = continent)) +
  # facetting by continent
  facet_wrap(~continent, scales = "free") +
  # make it a histogram
  geom_histogram(bins = 30)  +
  # make a better x axis
  scale_x_continuous(breaks = scales::pretty_breaks(n = 3))
```

```{r}
gapminder %>% 
  # continent as x axis, gdpPercap as y axis
  ggplot(aes(continent, gdpPercap)) +
  # scale y axis by log10
  scale_y_log10() +
  # make it a boxplot
  geom_boxplot()
```

A table provides accurate numbers on the spread of GDP per capita, but it cannot show the details inside the data. While histogram and boxplot provide better summaries (e.g. the boxplot contains outliers). Comparing histogram and boxplot, histogram only counts data within the same bins, while boxplot also shows some statistics facts (like quartiles) of data.

## Task #3

**Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.**

Let's calculate the weighted mean of life expectancy, weighting by population.

```{r}
means <- gapminder %>% 
  # group by year
  group_by(year) %>% 
  # calculate vanilla mean and weighted mean
  summarize(
    vanilla_mean_lifeExp = mean(lifeExp),
    weighted_mean_lifeExp_by_pop = weighted.mean(lifeExp, pop)
  )

# display as table
means %>% 
  knitr::kable()
```

Let's draw some line plots to show them.

```{r}
means %>% 
  ggplot() +
  # make a line plot with points for vanilla_mean_lifeExp
  geom_line(aes(year, vanilla_mean_lifeExp, color = "vanilla_mean_lifeExp")) +
  geom_point(aes(year, vanilla_mean_lifeExp, color = "vanilla_mean_lifeExp")) +
  # make a line plot with points for weighted_mean_lifeExp_by_pop
  geom_line(aes(year, weighted_mean_lifeExp_by_pop, color = "weighted_mean_lifeExp_by_pop")) +
  geom_point(aes(year, weighted_mean_lifeExp_by_pop, color = "weighted_mean_lifeExp_by_pop")) +
  # add legend for two lines
  scale_color_discrete("Mean type")
  
```

Basically, the life expectancy tends to increase thoughout the years.

## Task #4

**How is life expectancy changing over time on different continents?**

Let's first use a table to show the trend of how life expectancy changes. We also calculate the changes in every five years.

```{r}
changes <- gapminder %>% 
  # need to group by continent and year
  group_by(continent, year) %>% 
  # use mean as benchmark
  summarize(
    mean_lifeExp = mean(lifeExp)
  ) %>% 
  # calculate absolute change and relative change
  mutate(
    change_in_five_years = mean_lifeExp - lag(mean_lifeExp),
    accumulate_change = mean_lifeExp- first(mean_lifeExp)
  )

# display the table
changes %>% 
  knitr::kable()
```

Let's plot it out.

```{r}
changes %>% 
  # year as x axis and mean_lifeExp as y axis
  ggplot(aes(year, mean_lifeExp)) +
  # facetting by continent
  facet_wrap(~continent, scales="free_y")+
  # make a line plot with points
  geom_line() +
  geom_point()
```

```{r}
changes %>%
  # remove NA values
  filter(!is.na(change_in_five_years)) %>% 
  # year as x axis and change_in_five_years as y axis
  ggplot(aes(year, change_in_five_years)) +
  # facetting by continent
  facet_wrap(~continent, scales = "free_y")+
  # make a line plot with points
  geom_line() +
  geom_point()
```

```{r}
changes %>%
  # year as x axis and accumulate_change as y axis
  ggplot(aes(year, accumulate_change)) +
  # facetting by continent
  facet_wrap(~continent, scales = "free_y")+
  # make a line plot with points
  geom_line() +
  geom_point()
```

From the above two figures, we get to know:

- Except Africa suffered from a decrement between 1997 and 2002, all the continents have increment throughout the years.
- Generally speaking, except Oceania, the speeds of lifeExp grows of other continents become smaller and smaller.

## Task #5 

**Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy - you decide - a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.**

Let's first know the range and mean of worldwide life expectancy, so we can decide a benchmark.

```{r}
range(gapminder$lifeExp)
```

```{r}
mean(gapminder$lifeExp)
```

The mean is somehow near the middle of the range, so we can use it as a benchmark.

Let's first check if the lifeExp of a country in a certain year is higher/lower than the average.


```{r}
lifeExpCheck <- gapminder %>% 
  # group by country and year
  group_by(country, year) %>% 
  # check if a country has low lifeExp
  mutate(
    low_lifeExp = lifeExp < mean(gapminder$lifeExp)
  ) %>% 
  # select only necessary columns
  select(country, year, lifeExp, low_lifeExp)

# show previews
head(lifeExpCheck) %>% 
  knitr::kable()

tail(lifeExpCheck) %>% 
  knitr::kable()
```

To show the abundance of countries with low lifeExp, we can simply count the number of TRUE/FALSE in low_lifeExp. A bar plot is suitable in this case.

```{r}
lifeExpCheck %>% 
  # year as x axis, and show the portion of low_lifeExp
  ggplot(aes(year, fill = low_lifeExp)) +
  # make it a bar plot
  geom_bar()
```

From the figure, we know that the numbers of countries with low lifeExp become smaller and smller over the years.

## Task #6

**Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.**

Let's try to list data in countries in Americas first, and see if we can find something intersting.

```{r}
countries <- gapminder %>% 
  # filter continent as Americas
  filter(continent == "Americas") %>%
  # remove column continent
  select(-continent) %>% 
  # group by country and year
  group_by(country, year)

# show previews
head(countries) %>% 
  knitr::kable()

tail(countries) %>% 
  knitr::kable()
```

We try to explore the following things.

### Is there a relationship between lifeExp and pop?

```{r}
countries %>% 
  # pop as x axis and lifeExp as y axis
  ggplot(aes(pop, lifeExp)) +
  # scale y axis by log10
  scale_y_log10() +
  # facetting by country
  facet_wrap(~country, scales = "free") +
  # make a line plot
  geom_line() +
  # make a better x axis
  scale_x_continuous(breaks=NULL)
```

Interestingly, when the population gets larger, the life expectancy also becomes larger!

### What is the overall GDP of each country and how do they change over the years?

```{r}
countries_with_overall_gdp <- countries %>% 
  # calculate overall GDP
  mutate(
    overall_gdp = pop * gdpPercap
  )

countries_with_overall_gdp %>% 
  # year as x axis and overall_gdp as y axis
  ggplot(aes(year, overall_gdp)) +
  # facetting by country
  facet_wrap(~country, scales = "free") +
  # make a line plot
  geom_line() +
  # scale y axis by log10
  scale_y_log10() +
  # make a better x axis
  scale_x_continuous(breaks = scales::pretty_breaks(n = 2))
```

So we can notice some flactuations in these resultant lines. But it is a trend that every country has an increasing overall GDP.

### Does overall GDP affect life expectancy?

```{r}
countries_with_overall_gdp %>% 
  # overall_gdp as x axis and lifeExp as y axis
  ggplot(aes(overall_gdp, lifeExp)) +
  # facetting by country
  facet_wrap(~country, scales = "free") +
  # scale y axis by log10
  scale_y_log10() +
  # make it a line plot
  geom_line() +
  # make a better x axis
  scale_x_continuous(breaks = NULL)
```

Though we notice some flactuations in these lines, we somehow can conclude that when the overall GDP of a country is high, the lifeExp is also high.

# But I want to do more!

## Layout stretch goal: get table and figure side-by-side.

We directly use code form the third task, but put the table and figure side-by-side.

<style type="text/css">
.twoC {width: 100%}
.clearer {clear: both}
.twoC .table {max-width: 50%; float: left}
.twoC img {max-width: 50%; float: right}
</style>

<div class="twoC">

```{r results='asis', echo=FALSE}
gapminder %>% 
  # group by year
  group_by(year) %>% 
  # calculate vanilla mean and weighted mean
  summarize(
    vanilla_mean = mean(lifeExp),
    weighted_mean = weighted.mean(lifeExp, pop)
  ) %>% 
  # display as table
  knitr::kable()
```

```{r echo=FALSE}
means %>% 
  ggplot() +
  # make a line plot with points for vanilla_mean_lifeExp
  geom_line(aes(year, vanilla_mean_lifeExp, color = "vanilla_mean_lifeExp")) +
  geom_point(aes(year, vanilla_mean_lifeExp, color = "vanilla_mean_lifeExp")) +
  # make a line plot with points for weighted_mean_lifeExp_by_pop
  geom_line(aes(year, weighted_mean_lifeExp_by_pop, color = "weighted_mean_lifeExp_by_pop")) +
  geom_point(aes(year, weighted_mean_lifeExp_by_pop, color = "weighted_mean_lifeExp_by_pop")) +
  # add legend for two lines
  scale_color_discrete("Mean type")
```
</div>

<div class="clearer"></div>

## Table stretch goal: there are some really nice fancy table helper packages.

We show one example (not very fancy) using pander and xtable respectively, assuming these two packages have been already installed.

```{r}
suppressPackageStartupMessages(library(pander))
suppressPackageStartupMessages(library(xtable))
```


### pander

```{r}
emphasize.strong.cols(1)
emphasize.italics.cells(which(means >= 60, arr.ind = TRUE))
pander(means)
```


### xtable

```{r results='asis'}
print(xtable(means), type = "html")
```

# Notes
- Currently I do not know how to make better labels of x axis in facetting figures. Now I can only  suppress all of them for some figures in section "Task #6".
- I do not like the way to show a table and a figure side-by-side. The reason is, it can be only used in HTML, but on GitHub it does not support direct viewing of HTML.

# References
- [pander: An R Pandoc Writer](http://rapporter.github.io/pander/)
- [The xtable Gallery](https://cran.r-project.org/web/packages/xtable/vignettes/xtableGallery.pdf)